Climatic warming strengthens a positive feedback\\ between alpine shrubs and fire

### 2016-10-20 manuscript/manuscript.Rnw
Proof read edits

\begin{abstract}
Climate change is expected to increase fire activity and woody plant encroachment in arctic and alpine landscapes. However, the extent to which these increases interact to affect changes in structure, function and composition of alpine ecosystems is unknown. Here we use field surveys and experimental manipulations to examine warming and fire effects on recruitment, growth and survival of seedlings of four dominant, evergreen fire-killed alpine shrubs. We found that fire substantially increased shrub seedling establishment (up to \Sexpr{density_ratios(greaus_seedling_density_model, 'Grevillea')}\textendash fold) and that experimental warming doubled tall shrub seedling growth rates and could potentially increase survival. Warming had no effect on shrub recruitment, post-fire infilling of gaps by a dominant alpine grass, or competitive effects of grass on shrub seedling growth and survival. These findings indicate that warming, coupled with more frequent or severe fires, will likely result in an increase in the cover and abundance of evergreen shrubs by increasing recruitment, doubling growth rates, and potentially increasing shrub survival. Thus, warming is likely to strengthen an existing feedback between woody species abundance and fire activity in alpine and tundra environments.
\end{abstract}

@@

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\section*{Introduction}
Accurately forecasting the effects of climatic warming on vegetation dynamics requires an understanding of the mechanisms by which climate and vegetation interact. Most forecasting models include the direct effects of climatic conditions on species distributions, but largely ignore disturbances, particularly their type, frequency and severity \citep{Guisan:2005gz,Tucker:2012dr}. Recurrent disturbance is integral to all ecosystems, strongly influencing fundamental demographic processes such as recruitment and mortality, and thus the composition and structure of plant communities and biomes worldwide \citep{Bond:2005eg, Turner:2010wo}. Disturbance regimes are changing as a consequence of climate change \citep{Turner:2010wo, Westerling:2011jp,Bradstock:2014jp}. Moreover, the effects of disturbance will interact with climate change, and thus may amplify or diminish how changes to temperature and moisture affect vegetation \citep{Post:2008wz,Camac:2015hja,Enright:2015em}

Feedbacks between climate change, vegetation and fire will be particularly important determinants of the trajectory of ecosystem change in coming decades \citep{Bowman:2009kp}. For example, in addition to the direct effects of warming on fire weather, changing climate may also alter fire regimes by altering vegetation productivity, structure and plant trait composition \citep{Keeley:2012wo,Matthews:2012gia,Bowman:2014dr}. Feedbacks, which may be positive or negative, have been documented in tropical savanna \citep{Hoffmann:2003br}, boreal forests \citep{Krawchuk:2011wr}, Mediterranean ecosystems \citep{Keeley:2012wo}, grasslands \citep{Flannigan:2009kl} and alpine and arctic environments \citep{Goetz:2007gj,Wookey:2009wo}.

Alpine and arctic vegetation are particularly vulnerable to the effects of changing climate \citep{Engler:2011vq,Dullinger:2012cg, Elmendorf:2012kz}. Field manipulative experiments have been widely used to investigate the potential effects of climate change in alpine and tundra plant communities. These experiments have provided invaluable information on how climate directly influences plant phenology \citep{Hoffmann:2010kf, Dorji:2013tt}, reproduction \citep{Klady:2011dw}, morphology \citep{Hudson:2009cn}, growth \citep{Hollister:2005vf}, floristic composition \citep{Elmendorf:2012kz} and biotic interactions \citep{Klanderud:2005uw}.

However, most studies in arctic and alpine areas have focused on plant responses in mature, undisturbed vegetation \citep{Briceno:2015iv}. Few climate-change field experiments have included disturbance as a factor \citep[but see:][]{Munier:2010cg,Graae:2010ir,Camac:2015hja}, or examined the influence of climate change on seedlings in post-disturbance environments. Although alpine and tundra vegetation has a strong capacity to regenerate vegetatively \citep{Billings:1968bn} there is mounting evidence that seedling regeneration is important in these ecosystems \citep{Venn:2009bw, Briceno:2015iv}, particularly for woody species \citep{Camac:2013eb,Williams:2014ud}. Despite seedlings being highly vulnerable to climate and disturbance, they are the life stage that determines the long-term persistence of a species and its capacity to establish in new areas \citep{Walck:2011dm}. Little is known about how climate change affects seedling demography \citep{Briceno:2015iv}, particularly in alpine and tundra post-disturbance conditions. This is unfortunate, because if we wish to accurately predict future trajectories of vegetation change in these ecosystems, it is imperative we understand how seedlings respond to both changing climate and disturbance regimes \citep{Walck:2011dm, Briceno:2015iv}.

\section*{Conceptual feedback model}

In alpine and arctic ecosystems, warming experiments and long-term monitoring have documented significant increases in the growth and cover of woody species \citep{Sturm:2001vd,Walker:2006ht,MyersSmith:2011fp}. The frequency and extent of wildfires have also increased in these environments over recent decades, a trend expected to continue \citep{Westerling:2006ve, Flannigan:2009kl, Qiu:2009th, Bradstock:2014jp}. In alpine and tundra ecosystems, current evidence from long-term and experimental studies indicate that shrub recruitment and encroachment is highest in disturbed areas \citep{Williams:1987ws, Batllori:2009fe, Frost:2013tm} and lowest in areas with minimal bare ground cover, such as undisturbed alpine tussock grassland \citep{Williams:1987ws}. Current evidence indicates that climatic warming is likely to increase the growth rates of woody species \citep{Arft:1999iz, Elmendorf:2012ba, MyersSmith:2011fp} and that shrubs are the most flammable vegetation component of these ecosystems \citep{Williams:2006tl, Higuera:2009ga, Fraser:NHT5Kx_W}. Thus, increases in the cover of woody species in alpine and arctic environments is likely to increase the flammability of these ecosystems.

The results of these studies suggest a positive feedback between warmer temperatures, woody species and fire in alpine and tundra environments (Fig. \ref{fig1}a). Warmer temperatures are likely to lead to more frequent or severe fire. This is likely to increase recruitment opportunities for woody species and increase shrub thickening both within and beyond shrub boundaries \citep{Racine:2004tn,Lantz:2013fh}. If this effect is coupled with an increase in the growth and mortality of shrub seedlings, highly flammable fuels will accumulate faster, have higher landscape connectivity, and ultimately lead to increases in the likelihood of fire \citep{Higuera:2008dw, Higuera:2009ga,MyersSmith:2011fp}. Thus, warming could strengthen an existing climate-disturbance feedback, that not only has the potential to cause rapid changes in the composition and structure of alpine and arctic vegetation, but also has serious social, biodiversity and carbon sequestration consequences \citep{Mack:2011ew}.

@@

Here, we examine these unknowns and their impacts on this proposed feedback between climatic warming, shrubs and fire in Australian alpine vegetation. As with other tundra ecosystems, Australian alpine landscapes consist of a mosaic of vegetation types dominated by either herbaceous or woody species \citep{Williams:2014ud}. They have also experienced significant changes in climate. Since 1979, mean growing season temperatures have risen by approximately 0.4\textdegree C and annual precipitation has fallen by 6\% \citep{Wahren:2013ji}, with a consequent decline in snow pack depth \citep{SanchezBayo:2013ci}. These climatic changes have been correlated with a 10 to 20\% increase in shrub cover and a 25\% decline in graminoids cover \citep{Wahren:2013ji}. Much of the Australian Alps has also been burnt by recent (2003 and 2006) wildfires, the frequency and/or severity of which are expected to increase as a consequence of further climate change \citep{Bradstock:2014jp,Williams:2014ud}.

We took advantage of recent fires in alpine open heathland, a plant community that occupies ca. 25\% of the Australian alpine landscape \citep{Williams:2014ud}. Under global warming, this plant community is likely to encroach upon grasslands, wetlands and herbfields \citep[including the nationally endangered snow patch herbfields;][]{Williams:2015gj}, and is itself susceptible to shrub thickening \citep{McDougall:2003gb}. To identify some of the biotic and abiotic factors that affect shrub establishment and how warmer post-fire conditions affect shrub seedling vital rates, we combined field observations of post-fire seedling densities with a warming experiment that used seedlings of four dominant Australian evergreen fire-killed shrubs: \textit{Grevillea australis} (Proteaceae; a tall shrub), \textit{Asterolasia trymalioides} (Rutaceae; a prostrate shrub), \textit{Phebalium squamulosum} (Rutaceae; a tall shrub) and \textit{Prostanthera cuneata} (Lamiaceae; a tall shrub).

@@

The Australian Alps are an ideal place to examine potential fire-climate-vegetation feedbacks in high mountain environments because they have all the elements needed to demonstrate such potential feedbacks. The Alps are subject to recurrent landscape fire, and were burnt extensively in 2003 \citep{Williams:2006tl,Williams:2014ud}. The vegetation consists of a range of life forms (graminoids, other herbs, shrubs), the abundance of which strongly determines landscape flammability \citep{Williams:2006tl,Fraser:NHT5Kx_W}. There is also evidence that both fire regimes and the relative abundance of life forms in the Australian Alps have changed as a consequence of recent climate change \citep{Bradstock:2014jp,Wahren:2013ji,Camac:2015hja}.

We investigated shrub-grass dynamics in open heathland vegetation on the Bogong High Plains, Australia. Open heathland is a common Australian alpine vegetation type, occupying approximately 25\% of the treeless landscapes above 1600m in the Australian Alps. It is dominated by the shrub \textit{Grevillea australis} (shrub cover generally 20\textendash 60\%) with inter-shrub regions occupied by snow grasses (\textit{Poa hiemata} and \textit{Poa costiniana}) and other herbaceous species. It is an ecotone between closed heathland (\textgreater 70\% shrub cover) and tussock grassland (shrub cover < 20\%). \textit{Grevillea australis} is an obligate seeding shrub, and establishment of seedlings is dependent on disturbance creating bare ground in the grass\textendash herb sward \citep{Williams:1987ws}. Relative to tussock grassland, open heathland is a highly flammable plant community \citep{Williams:2006tl,Fraser:NHT5Kx_W}. A consequence of this differential flammability led to the 2003 wildfires on the Bogong High Plains burning approximately 60\% of open heathland and only 13\% of tussock grasslands. Following these fires, seedling regeneration of \textit{Grevillea australis} and other shrubs was prolific \citep{Williams:2014ud}.

@@

\section*{Discussion}
\subsection*{Strengthening of the warming-shrub-fire feedback}

In this study we have quantified several unknown interactions between fire, shrub-grass relationships and climate to extend a conceptual model of alpine shrub dynamics under climate change (Fig. \ref{fig1}b). In particular, we have shown how these interactions strengthen a hypothesized feedback loop that can rapidly increase shrub cover in the Australian Alps, and potentially other alpine and tundra ecosystems \citep{Higuera:2008dw}. While high cover of tussock grass is known to inhibit the establishment of shrub seedlings \citep{Williams:1987ws}, more frequent and severe fire caused by changing climate, will create bare ground required for shrub seedlings to establish. Our results highlight that if a shrub seedling establishes in bare ground, its growth and survival rates are not affected by proximity to tussock grass, irrespective or warming. More importantly, our results indicate that for tall shrubs, a 1\textdegree C increase in mean growing season temperature resulted in a doubling of growth rates and a potential increase in survival. These demographic effects will manifest in shrub thickening within and at shrub community boundaries, and ultimately, facilitate shrub establishment into non-shrubby communities.

Evidence from field studies of burning patterns \citep{Williams:2006tl}, quantification of fuel mass and architecture \citep{Fraser:NHT5Kx_W} and paleoecological studies \citep{Higuera:2008dw} indicate that shrubs are the most flammable component of alpine and tundra ecosystems \citep[significantly more than tussock grassland; ][]{Fraser:NHT5Kx_W}. Consequently, our results indicate that flammable fuel loads will accumulate twice as fast under a warmer climate. This is likely to further strengthen the feedback \citep{Higuera:2008dw, Higuera:2009ga} by increasing the frequency and severity of fires, which then creates more shrub recruitment opportunities both within and beyond shrub-grassland boundaries. Such changes may persist for decades.

While we have addressed several unknowns associated with this feedback between climatic warming, shrubs and fire, there are others we have not addressed that may also strengthen, weaken or break this feedback. The most obvious mechanism that will break this cycle involves shorter fire intervals that prevent fire-killed shrubs reaching reproductive age and thereby exhausting the seedbank \citep{Enright:2015em}. However, this scenario is unlikely for the majority of alpine (or tundra) landscapes, including those in Australia, where current fire intervals of 50 to 100 years would need to decrease to less than 20 years\textemdash the time estimated for the species in this study to reach reproductive maturity \citep{Williams:2008ba}. Furthermore, if, as hypothesised, reproductive output is related to plant size \citep{Wenk:2015ko}, then climatic warming may allow fire-killed species to reach reproductive maturity sooner, and consequently, may increase species resilience to short fire intervals. Nevertheless, these factors and others such as changes in snow pack \citep{Wipf:2009jh}, soil moisture \citep{Engler:2011vq}, herbivory \citep{Post:2008wz} and adaptation \citep{Byars:2009fd, Hoffmann:2011fp} require further research because they are likely to be altered by the interactive effects of climate and disturbance in unpredictable ways \citep{MyersSmith:2011fp}.

By focusing on seedlings, the life stage most vulnerable to climate and disturbance, and which determines the long-term persistence of shrub species and its capacity to establish in new areas \citep{Walck:2011dm, Briceno:2015iv}, our findings provide a possible explanation as to why shrub cover has increased in the Australian Alps, often at the expense of grasslands \citep{McDougall:2003gb,Wahren:2013ji}. An explanation that may also apply to other arctic and alpine ecosystems \citep{Racine:2004tn,MyersSmith:2011fp,Frost:2013tm,Lantz:2013fh}. But more importantly, our results provide evidence for the underlying processes that could result in a warming-fire-shrub feedback that has been hypothesized in paleoecological studies \citep{Higuera:2008dw, Higuera:2009ga}. Based on current observations and predictions, average global temperature has already increased by 0.85\textdegree C since 1880 and is expected to rise by as much as 4.8\textdegree C by 2100 \citep{IPCC:2013vv}. In alpine and tundra environments, temperatures \citep{ChapinIII:2005js}, shrub cover \citep{MyersSmith:2011fp} and the frequency and severity of fire \citep{Westerling:2006ve, Flannigan:2009kl,Qiu:2009th, Bradstock:2014jp} have all increased across the globe in the last few decades as a consequence of rising temperatures. These changes mean that the warming-shrub-fire feedback loop identified here is likely to have already strengthened. Thus, other non-woody communities will become shrubbier and more prone to fire, the effects of which, will have significant consequences for carbon sequestration, water supply and biodiversity.

>>>>

\begin{abstract}
Climate change is expected to increase fire and woody plant encroachment in arctic and alpine landscapes. However, the extent to which these increases interact to affect the structure, function and composition of alpine ecosystems is largely unknown. Here we use field surveys and experimental manipulations to examine how warming and fire affect recruitment, seedling growth and seedling survival in four dominant Australian alpine shrubs. We found that fire increased establishment of shrub seedlings by as much as \Sexpr{density_ratios(greaus_seedling_density_model, 'Grevillea')}\textendash fold. Experimental warming also doubled growth rates of tall shrub seedlings and could potentially increase their survival. By contrast, warming had no effect on shrub recruitment, post-fire tussock regeneration, or how tussock grass affected shrub seedling growth and survival. These findings indicate that warming, coupled with more frequent or severe fires, will likely result in an increase in the cover and abundance of evergreen shrubs. Given shrubs are one of the most flammable components in alpine and tundra environments, warming is likely to strengthen an existing feedback between woody species abundance and fire in these ecosystems.
\end{abstract}

@@

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\section*{Introduction}
Accurately forecasting the effects of climatic warming on vegetation dynamics requires an understanding of the mechanisms by which climate and vegetation interact. Most forecasting models include the direct effects of climatic conditions on species distributions, but largely ignore disturbances, particularly their type, frequency and severity \citep{Guisan:2005gz,Tucker:2012dr}. Recurrent disturbance is integral to all ecosystems, strongly influencing fundamental demographic processes such as recruitment and mortality, and thus the composition and structure of plant communities and biomes worldwide \citep{Bond:2005eg, Turner:2010wo}. Disturbance regimes are already changing as a consequence of climate change \citep{Turner:2010wo, Westerling:2011jp,Bradstock:2014jp}. It is therefore imperative we understand how the effects of disturbance will interact with climate change, and whether such effects amplify or diminish how vegetation responds to changing temperature and moisture \citep{Post:2008wz,Camac:2015hja,Enright:2015em}

Understanding the feedback between climate, vegetation and fire will be particularly important in order to accurately predict the trajectory of ecosystem change in coming decades \citep{Bowman:2009kp}. In addition to the direct effects of warming on fire weather, changing climate may also indirectly alter fire regimes by altering vegetation productivity, structure and composition \citep{Keeley:2012wo,Matthews:2012gia,Bowman:2014dr}. Such feedbacks may be positive or negative, and have been documented in tropical savanna \citep{Hoffmann:2003br}, boreal forests \citep{Krawchuk:2011wr}, grasslands \citep{Flannigan:2009kl} and alpine and arctic environments \citep{Goetz:2007gj,Wookey:2009wo}.

Alpine and arctic vegetation are considered to be particularly vulnerable to the effects of changing climate \citep{Engler:2011vq,Dullinger:2012cg, Elmendorf:2012kz}. In these ecosystems, field manipulative experiments have shown that climate directly influences plant phenology \citep{Hoffmann:2010kf, Dorji:2013tt}, reproduction \citep{Klady:2011dw}, morphology \citep{Hudson:2009cn}, growth \citep{Hollister:2005vf}, floristic composition \citep{Elmendorf:2012kz} and biotic interactions \citep{Klanderud:2005uw}. However, most studies in these ecosystems have focused on mature plant responses in undisturbed vegetation \citep{Briceno:2015iv}. Few have included disturbance as a factor \citep[but see:][]{Munier:2010cg,Graae:2010ir,Camac:2015hja}, or examined the influence of climate change on seedlings in post-disturbance environments. As a consequence, little is known about how climate affect seedling demographic rates \citep{Briceno:2015iv}, particularly in alpine and tundra post-disturbance conditions. This is despite mounting evidence that seedling regeneration is important in alpine and arctic ecosystems \citep{Venn:2009bw, Briceno:2015iv}, particularly for woody species \citep{Camac:2013eb,Williams:2014ud}. Seedlings are the life stage that determines the long-term persistence of a species as well as its capacity to establish in new areas \citep{Walck:2011dm}. As such, in order to accurately predict future trajectories of vegetation change in alpine and arctic ecosystems, it is impertative we understand how seedlings respond to both changing climate and disturbance regimes \citep{Walck:2011dm, Briceno:2015iv}.

\section*{Feedback between climatic warming, fire and shrubs}

In alpine and arctic ecosystems, warming experiments and long-term monitoring have documented significant increases in the growth and cover of woody species \citep{Sturm:2001vd,Walker:2006ht,MyersSmith:2011fp}. The frequency and extent of wildfires in these environments have also increased over recent decades, a trend expected to continue \citep{Westerling:2006ve, Flannigan:2009kl, Qiu:2009th, Bradstock:2014jp}. Current evidence from these ecosystems indicate that shrub recruitment and encroachment is highest in disturbed areas and lowest in areas with minimal bare ground cover \citep{Williams:1987ws, Batllori:2009fe, Frost:2013tm}. Evidence also suggests that climatic warming is likely to increase growth rates of woody species \citep{Arft:1999iz, Elmendorf:2012ba, MyersSmith:2011fp} and that shrubs are potentially the most flammable vegetation component in these ecosystems \citep{Williams:2006tl, Higuera:2009ga, Fraser:NHT5Kx_W}. Thus, increases in the cover of woody species in alpine and arctic environments may likely to increase the flammability of these ecosystems.

The results of these studies suggest a positive feedback could exist between warmer temperatures, woody species and fire in alpine and tundra environments (Fig. \ref{fig1}a). Specifically, warmer temperatures may lead to more frequent and severe fire, which in turn, may increase recruitment opportunities (i.e. more bare ground) for woody species and increase shrub thickening both within and beyond shrub boundaries \citep{Racine:2004tn,Lantz:2013fh}. If this effect is coupled with an increase in the growth and survival of shrub seedlings, highly flammable fuels will accumulate at a faster rate, have higher landscape connectivity, and ultimately lead to increases in the likelihood of fire. The consequence of which will further increase shrub recruitment opportunities. Thus, warming could strengthen an existing climate-disturbance feedback, that not only has the potential to cause rapid changes in the composition and structure of alpine and arctic vegetation, but also has serious social, biodiversity and carbon sequestration consequences \citep{Mack:2011ew}.

@@

Here, we examine these unknowns and their impacts on this proposed feedback between climatic warming, shrubs and fire in Australian alpine vegetation. As with other tundra ecosystems, Australian alpine landscapes consist of a mosaic of vegetation types dominated by either herbaceous or woody species \citep{Williams:2014ud}. They have also experienced significant changes in climate. Since 1979, mean growing season temperatures have risen by approximately 0.4\textdegree C and annual precipitation has fallen by 6\% \citep{Wahren:2013ji}, with a consequent decline in snow pack depth \citep{SanchezBayo:2013ci}. These climatic changes have been correlated with a 10 to 20\% increase in shrub cover and a 25\% decline in graminoids cover \citep{Wahren:2013ji}. Much of the Australian Alps has also been burnt by recent (2003 and 2006) wildfires, the frequency and severity of which are expected to increase as a consequence of further climate change \citep{Bradstock:2014jp,Williams:2014ud}.

In this study we took advantage of recent fires in alpine open heathland, a plant community that occupies ca. 25\% of the Australian alpine landscape \citep{Williams:2014ud}. Under global warming, open heathland is likely to encroach upon grasslands, wetlands and herbfields \citep[including the nationally endangered snow patch herbfields;][]{Williams:2015gj}, and is itself susceptible to shrub thickening \citep{McDougall:2003gb}. To identify some of the biotic and abiotic factors that affect shrub establishment and how warmer post-fire conditions affect shrub seedling vital rates, we combined field observations of post-fire seedling densities with a warming experiment that used seedlings of four dominant Australian evergreen fire-killed shrubs: \textit{Grevillea australis} (Proteaceae; a tall shrub), \textit{Asterolasia trymalioides} (Rutaceae; a prostrate shrub), \textit{Phebalium squamulosum} (Rutaceae; a tall shrub) and \textit{Prostanthera cuneata} (Lamiaceae; a tall shrub).

@@

The Australian Alps are an ideal place to examine the proposed fire-climate-vegetation feedback because they have all the elements needed to demonstrate such a feedback. The Alps are subject to recurrent landscape fire (approximately every 50 to 100 years), and were burnt extensively in 2003 \citep{Williams:2006tl,Williams:2014ud}. The vegetation consists of a range of life forms (graminoids, forbs and shrubs), of which shrub abundance strongly determines landscape flammability \citep{Williams:2006tl,Fraser:NHT5Kx_W}. There is also evidence that both fire regimes \citep{Bradstock:2014jp} and the relative abundance of life forms in the Australian Alps have changed as a consequence of recent climate change \citep{Wahren:2013ji,Camac:2015hja}.

We investigated shrub-grass dynamics in open heathland on the Bogong High Plains, Australia. Open heathland is a common Australian alpine vegetation type, occupying approximately 25\% of the treeless landscapes above 1600m in the Australian Alps. It is an ecotone between closed heathland (\textgreater 70\% shrub cover) and tussock grassland (shrub cover < 20\%). Open heathland is dominated by the shrub \textit{Grevillea australis} (shrub cover generally 20\textendash 60\%) with inter-shrub regions occupied by snow grasses (\textit{Poa hiemata} and \textit{Poa costiniana}) and other herbaceous species. \textit{Grevillea australis} is an obligate seeding shrub, and establishment of seedlings is dependent on disturbance creating bare ground in the grass\textendash herb sward \citep{Williams:1987ws}. Relative to tussock grassland, open heathland is a highly flammable plant community \citep{Williams:2006tl,Fraser:NHT5Kx_W}. A consequence of this differential flammability led to the 2003 wildfires on the Bogong High Plains burning approximately 60\% of open heathland and only 13\% of tussock grasslands. Following these fires, seedling regeneration of \textit{Grevillea australis} and other shrubs was prolific \citep{Williams:2014ud}.

@@

\section*{Discussion}

In this study we have quantified several unknown interactions between fire, shrub-grass relationships and climate to extend a conceptual model of alpine shrub dynamics under climate change (Fig. \ref{fig1}b). In particular, we have shown how these interactions strengthen a hypothesized feedback that can rapidly increase shrub cover in the Australian Alps, and potentially, other alpine and tundra ecosystems \citep{Higuera:2008dw}. While high cover of tussock grass is known to inhibit the establishment of shrub seedlings \citep{Williams:1987ws}, fire can create bare ground neccesary for shrub seedling establishment. Our results highlight that fire can increase shrub seedling establishment by as much \Sexpr{density_ratios(greaus_seedling_density_model, 'Grevillea')}\textendash fold. Our results also indicate that if a shrub seedling manages to establish in a bare ground patch, its rate of growth and survival are not affected by tussock grass proximity, irrespective or warming. More importantly, our findings indicate that for tall shrubs, a 1\textdegree C increase in mean growing season temperature will result in a doubling of growth rates and a potential increase in survival. The consequence of these demographic effects will likely manifest as shrub thickening within and at shrub community boundaries. This in turn, will facilitate shrub establishment, via increased seed pools, into non-shrubby communities.

Landscape flammability, and thus fire, are also likely to increase as a result of warming effects on shrub seedling vital rates. Evidence from field studies of burning patterns \citep{Williams:2006tl}, quantification of fuel mass and architecture \citep{Fraser:NHT5Kx_W} and paleoecological studies \citep{Higuera:2008dw, Higuera:2009ga} indicate that shrubs are the most flammable component of alpine and tundra ecosystems \citep[significantly more than tussock grassland; ][]{Fraser:NHT5Kx_W}. In the Australian Alps, this differential flammability resulted in approximately 60\% of open heathland and only 13\% of tussock grasslands being burnt by the extensive 2003 wildfires. Consequently, our results indicate that flammable fuel loads will accumulate twice as fast under a warmer climate. Ultimately this will further strengthen the feedback between shrubs and fire by increasing the frequency and severity of fires, which in turn, will create more bare ground, and thus more shrub recruitment opportunities both within and beyond current shrub boundaries. Recruitment opportunities that could persist for decades \citep{Williams:2014ud}.

While we have addressed several unknowns associated with this feedback between climatic warming, shrubs and fire, there are others we have not addressed that may also strengthen, weaken or break this feedback. The most obvious mechanism that will break this cycle involves short fire intervals that prevent fire-killed shrubs reaching reproductive age and thereby exhausting the seedbank \citep{Enright:2015em}. However, this scenario is unlikely for the majority of alpine (or tundra) landscapes, including those in Australia. For example, in the Australian Alps, current fire intervals of 50 to 100 years would need to decrease to less than 20 years\textemdash the time estimated for the species in this study to reach reproductive maturity \citep{Williams:2008ba}. Furthermore, if reproductive output is related to plant size \citep{Wenk:2015ko}, then climatic warming may allow fire-killed species to reach reproductive maturity sooner, and consequently, may increase their resilience to short fire intervals. Nevertheless, this and other factors such as changes in snow pack \citep{Wipf:2009jh}, soil moisture \citep{Engler:2011vq}, herbivory \citep{Post:2008wz} and adaptation \citep{Byars:2009fd, Hoffmann:2011fp} require further research because they are all likely to be altered by the interactive effects of climate and disturbance in unpredictable ways.

By focusing on the life stage most vulnerable to climate and disturbance, and which determines a species capacity to establish in new areas \citep{Walck:2011dm, Briceno:2015iv}, our analyses provide a possible explanations as to why shrub cover has increased in the Australian Alps, often at the expense of grasslands \citep{McDougall:2003gb,Wahren:2013ji}. Explanations that may also apply to other arctic and alpine ecosystems \citep{Racine:2004tn,MyersSmith:2011fp,Frost:2013tm,Lantz:2013fh}. But more importantly, our results provide evidence for the underlying processes that could result in a warming-fire-shrub feedback that has been hypothesized in arctic paleoecological studies \citep{Higuera:2008dw, Higuera:2009ga}. Based on current observations, average global temperature has already increased by 0.85\textdegree C since 1880 and is expected to rise by as much as 4.8\textdegree C by 2100 \citep{IPCC:2013vv}. In alpine and tundra environments, temperatures \citep{ChapinIII:2005js}, shrub cover \citep{MyersSmith:2011fp} and the frequency and severity of fire \citep{Westerling:2006ve, Flannigan:2009kl,Qiu:2009th, Bradstock:2014jp} have all increased in the last few decades. These changes mean that the warming-shrub-fire feedback identified here is likely to have already strengthened. If this is the case, other non-woody communities will become shrubbier and more flammable, the effects of which, will have significant consequences for carbon sequestration, water supply and biodiversity.
